---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pdokr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/pdokr)](https://CRAN.R-project.org/package=pdokr)
[![R-CMD-check](https://github.com/coeneisma/pdokr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coeneisma/pdokr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`pdokr` makes it easy to work with open geographic data from
[PDOK](https://www.pdok.nl/) (Publieke Dienstverlening Op de Kaart), the
national geodata platform of the Netherlands. It helps you discover datasets and
their layers, load a layer as an [`sf`](https://r-spatial.github.io/sf/) object
(with automatic pagination and explicit coordinate reference system handling),
filter data by any polygon area, and geocode addresses and place names.

## Installation

You can install the development version of `pdokr` from
[GitHub](https://github.com/coeneisma/pdokr) with:

``` r
# install.packages("devtools")
devtools::install_github("coeneisma/pdokr")
```

## Example

```{r example}
library(pdokr)

# Search for a dataset
pdok_search_datasets("nationale parken")

# Load a layer as an sf object
parks <- pdok_read("rvo/nationale-parken-geharmoniseerd", "protectedsite")

# Filter to an area: the province of Utrecht, found by geocoding
utrecht <- pdok_geocode("Utrecht", type = "provincie")
pdok_filter_by(parks, utrecht)
```

## Learn more

See the [package website](https://coeneisma.github.io/pdokr/) for the reference
documentation and articles on getting started, filtering data by area, working
with coordinate reference systems, and querying PDOK by hand.
