---
title: "3D buildings from BAG and AHN"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

[Mapping buildings by construction year](bag-buildings.html) drew building
footprints flat. Here we lift them into **3D**, using two PDOK sources together:
the footprints from the BAG (vector, `sf`) and the heights from AHN (raster,
`terra`). The recipe is simple — a footprint plus a height is an extruded block —
and it is built entirely from open PDOK data. We take the cluster of towers
around Rotterdam's Erasmusbrug, a real skyline.

```{r setup}
library(pdokr)
library(sf)
library(terra)
library(mapgl)
```

## An area, from a place name

We geocode the Erasmusbrug — the bridge that spans the river — and take a 750 m
circle around its centre, which covers the towers on both banks (Kop van Zuid,
the Boompjes, the Zalmhaventoren). Asking for RD New coordinates (EPSG:28992)
means the footprints and the AHN raster share one coordinate system, so no
reprojection is needed later.

```{r area}
bridge <- pdok_geocode("Erasmusbrug, Rotterdam", type = "weg", crs = 28992)
area <- st_buffer(st_centroid(st_geometry(bridge)), 750)
```

## Footprints from the BAG

```{r footprints}
footprints <- pdok_read("kadaster/bag", "pand", filter_by = area, crs = 28992)
nrow(footprints)
```

## Heights from AHN

AHN comes in two flavours: the **DSM** (digital surface model — the top of
everything, including roofs) and the **DTM** (digital terrain model — the bare
ground). Their difference is the height of whatever stands on the ground.

```{r heights}
object_height <- pdok_ahn(area, "dsm") - pdok_ahn(area, "dtm")
names(object_height) <- "height"
```

`terra::extract()` reads that height onto each footprint. We take the *highest*
cell within each polygon — the top of the tower rather than its podium:

```{r extract}
footprints$height <- terra::extract(
  object_height, footprints, fun = max, na.rm = TRUE
)$height

# keep the buildings with a sensible height, and only the columns we map
buildings <- footprints[
  is.finite(footprints$height) & footprints$height > 2,
  c("identificatie", "height")
]
range(round(buildings$height))
```

## Extrude into 3D

A `maplibre` map extrudes the footprints by the height column. On the PDOK vector
basemap, tilted, Rotterdam's river towers rise over the Maas — drag to look
around, and click a building for its height.

```{r map-3d}
maplibre(
  style = pdok_basemap("standaard", format = "vector"),
  pitch = 55, projection = "mercator"
) |>
  add_fill_extrusion_layer(
    id = "buildings", source = buildings,
    fill_extrusion_color = "#E8631C",
    fill_extrusion_height = get_column("height"),
    fill_extrusion_opacity = 0.9,
    tooltip = "height"
  ) |>
  add_control(html = "Kaartgegevens &copy; Kadaster", position = "bottom-right") |>
  fit_bounds(buildings, animate = FALSE)
```

That is real PDOK data in 3D: BAG footprints, AHN heights, combined in R.

PDOK also publishes a ready-made 3D city model — the *3D Basisvoorziening* — as
OGC API **3D Tiles**. Those you would view with a 3D-tiles renderer such as
[`rdeck`](https://github.com/anthonynorth/rdeck) (deck.gl's `Tile3DLayer`),
pointing it at the tileset URL; that is rendering a finished model, rather than
the `sf` + `terra` analysis we did here.

## Where to next

* [Features and rasters](features-and-rasters.html) — the vector/raster
  distinction this builds on.
* [PDOK basemaps](basemaps.html) — the vector basemap used here.
