---
title: "Mapping buildings by construction year (BAG)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

The [BAG](https://www.pdok.nl/) (Basisregistratie Adressen en Gebouwen) holds
every building in the Netherlands, with attributes such as the year it was
built. It is served over the OGC API Features service, so `pdok_read()` reads it
like any other layer. This article maps the buildings of one neighbourhood,
coloured by construction year.

```{r setup}
library(pdokr)
library(tmap)
library(dplyr)
```

## Choose an area

We use the administrative boundaries to select the `Stadsdriehoek`, the historic
centre of Rotterdam.

```{r area}
gemeenten <- pdok_read(
  "cbs/gebiedsindelingen", "gemeente_gegeneraliseerd", datetime = 2025
)
rotterdam <- filter(gemeenten, statnaam == "Rotterdam")

buurten <- pdok_read(
  "cbs/gebiedsindelingen", "buurt_gegeneraliseerd", datetime = 2025,
  filter_by = rotterdam, predicate = "within"
)
stadsdriehoek <- filter(buurten, statnaam == "Stadsdriehoek")
```

## Load every building in it

The `kadaster/bag` dataset has a `pand` (building) layer. We filter it to the
neighbourhood; `pdok_read()` pre-filters at the server and clips to the exact
shape.

```{r buildings}
buildings <- pdok_read("kadaster/bag", "pand", filter_by = stadsdriehoek)
nrow(buildings)
```

BAG records an unknown construction year as `9999`; we set those to `NA`.

```{r clean}
buildings <- mutate(buildings, bouwjaar = na_if(bouwjaar, 9999))
range(buildings$bouwjaar, na.rm = TRUE)
```

## Map by construction year

Rotterdam's historic centre was largely destroyed in the May 1940 bombing and
rebuilt afterwards. To make that dividing line jump out, we colour the buildings
that pre-date the bombing (before 1940) in **cool blues** and the post-war
rebuild in **warm tones** — the surviving pre-war buildings stand out against a
sea of reconstruction.

```{r map-static}
tmap_mode("plot")

era_scale <- tm_scale_intervals(
  breaks = c(1600, 1900, 1940, 1960, 1980, 2000, 2026),
  values = c("#1f78b4", "#a6cee3", "#ffffb2", "#fecc5c", "#fd8d3c", "#e31a1c")
)

tm_basemap(pdok_basemap("grijs")) +
  tm_shape(buildings) +
  tm_polygons(fill = "bouwjaar", fill.scale = era_scale, col = NULL) +
  tm_title("Buildings in Rotterdam's Stadsdriehoek, by construction year") +
  tm_credits("Kaartgegevens © Kadaster")
```

The same map interactively — zoom in and click a building:

```{r map-interactive}
tmap_mode("view")

tm_basemap(pdok_basemap("grijs")) +
  tm_shape(buildings) +
  tm_polygons(fill = "bouwjaar", fill.scale = era_scale, col = NULL,
              id = "identificatie", popup = tm_popup(vars = c("Built" = "bouwjaar"))) +
  tm_credits("Kaartgegevens © Kadaster")
```

```{r reset, include = FALSE}
tmap_mode("plot")
```

## Where to next

* [Filtering data by area](filtering-by-area.html) — the load-then-filter
  workflow used here.
* [Combining with external data](duo-schools.html) — join `pdokr` with another
  open API.
* [PDOK basemaps](basemaps.html) — the grey background map used here, and the
  other styles.
