---
title: "PDOK basemaps"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

`pdokr` is for reading PDOK *data*. For the *background* of a map — the reference
topography under your own layers — PDOK publishes its official basiskaart, and
`pdok_basemap()` hands you the URL for it. Nothing is downloaded: you get a URL
to pass to your mapping package, so it is instant and works offline.

```{r setup}
library(pdokr)
library(tmap)
library(sf)
library(dplyr)
library(mapgl)
library(leaflet)
```

## Two kinds of basemap

`pdok_basemap()` returns one of two things, set by `format`:

- **raster** (the default) — a classic tile URL (`{z}/{x}/{y}`, PNG/JPEG). Works
  with `tmap`, `leaflet` *and* `maplibre`/`mapgl`. The safe, universal choice.
- **vector** — a Mapbox-GL *style* URL. Works with `maplibre`/`mapgl` only, but
  stays razor-sharp at every zoom and can tilt into 3D.

```{r show-urls}
pdok_basemap()                                # raster, default style
pdok_basemap("grijs")                         # another raster style
pdok_basemap("standaard", format = "vector")  # a vector style URL
```

The available styles differ per format:

| `format = "raster"` | `format = "vector"` |
|---|---|
| `standaard`, `grijs`, `pastel`, `water`, `luchtfoto` | `standaard`, `zonder_labels`, `luchtfoto`, `darkmode` |

(`luchtfoto` is aerial imagery.)

## Raster basemaps — tmap & leaflet

Hand the URL to `tm_basemap()`. Here are the twelve province outlines on the
standard PDOK basemap — drawing only the borders keeps the basemap readable:

```{r data}
provinces <- pdok_read(
  "cbs/gebiedsindelingen", "provincie_gegeneraliseerd", datetime = 2025
)
```

```{r raster-tmap}
tmap_mode("view")

tm_basemap(pdok_basemap("standaard")) +
  tm_shape(provinces) +
  tm_borders(col = "#E8631C", lwd = 1.5) +
  tm_credits("Kaartgegevens © Kadaster")
```

```{r reset, include = FALSE}
tmap_mode("plot")
```

In plain `leaflet` the same URL is the `urlTemplate`:

```{r leaflet, eval = FALSE}
leaflet() |>
  addTiles(
    urlTemplate = pdok_basemap("standaard"),
    attribution = "Kaartgegevens &copy; Kadaster"
  ) |>
  addPolygons(data = st_transform(provinces, 4326))
```

A raster basemap is ordinary web tiles, so it also works in **static** maps:
switch to `tmap_mode("plot")` (or use `ggplot2` with `maptiles`/`ggspatial`) and
the tiles are downloaded and drawn into the image. The vector style, by contrast,
is rendered live in the browser, so it is interactive-only.

```{r raster-static}
tmap_mode("plot")

utrecht <- filter(provinces, statnaam == "Utrecht")

tm_basemap(pdok_basemap("standaard")) +
  tm_shape(utrecht) +
  tm_borders(col = "#E8631C", lwd = 2) +
  tm_credits("Kaartgegevens © Kadaster")
```

## The styles, side by side {.tabset}

The five raster styles over the same spot in Utrecht. Each tab is just
`leaflet() |> addTiles(pdok_basemap("<style>"))` — only the style name changes.

### standaard

```{r gal-standaard, echo = FALSE}
leaflet() |>
  addTiles(pdok_basemap("standaard"), attribution = "&copy; Kadaster") |>
  setView(5.121, 52.090, zoom = 14)
```

### grijs

```{r gal-grijs, echo = FALSE}
leaflet() |>
  addTiles(pdok_basemap("grijs"), attribution = "&copy; Kadaster") |>
  setView(5.121, 52.090, zoom = 14)
```

### pastel

```{r gal-pastel, echo = FALSE}
leaflet() |>
  addTiles(pdok_basemap("pastel"), attribution = "&copy; Kadaster") |>
  setView(5.121, 52.090, zoom = 14)
```

### water

```{r gal-water, echo = FALSE}
leaflet() |>
  addTiles(pdok_basemap("water"), attribution = "&copy; Kadaster") |>
  setView(5.121, 52.090, zoom = 14)
```

### luchtfoto

```{r gal-luchtfoto, echo = FALSE}
leaflet() |>
  addTiles(pdok_basemap("luchtfoto"), attribution = "&copy; Kadaster") |>
  setView(5.121, 52.090, zoom = 14)
```

## Vector basemaps — maplibre / mapgl

[`mapgl`](https://walker-data.com/mapgl/) brings the MapLibre and Mapbox GL
libraries to R: vector maps that stay sharp at any zoom and can tilt into 3D. A
GL map takes a *style* URL — exactly what `pdok_basemap(format = "vector")`
returns.

```{r vector-maplibre}
maplibre(style = pdok_basemap("standaard", format = "vector")) |>
  add_line_layer(
    id = "provinces", source = provinces,
    line_color = "#E8631C", line_width = 2
  ) |>
  fit_bounds(provinces, animate = FALSE)
```

## Which should I use?

| | raster | vector |
|---|---|---|
| Works in | tmap, leaflet, mapgl | maplibre / mapgl |
| Extra package | none (you likely have tmap or leaflet) | `mapgl` |
| Look | fixed tiles, fine for most maps | razor-sharp, tiltable, 3D-capable |
| Best for | quick maps, the common case | polished, interactive, 3D |

When in doubt, use the **raster** default — it works everywhere.

## Attribution

The basemap data is © Kadaster / PDOK. Show this on any map that uses it — via
`tm_credits()` in `tmap`, the `attribution =` argument in `leaflet`. The vector
style already carries its attribution, which `maplibre`/`mapgl` displays.

## Where to next

* [Getting started](getting-started.html) — reading and mapping PDOK data.
* [Filtering data by area](filtering-by-area.html) — `pdok_filter_by()`.
