---
title: "Thematic maps from CBS statistics"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

Statistics Netherlands ([CBS](https://www.cbs.nl/)) publishes a yearly dataset
with hundreds of figures per neighborhood — population, income, housing, energy
use, distance to amenities, and much more. Because PDOK serves it with geometry,
a single `pdok_read()` gives you everything you need for a choropleth. This
article maps the population density across the neighborhoods of Amsterdam.

```{r setup}
library(pdokr)
library(tmap)
library(dplyr)
```

## Read the neighborhoods of one municipality

The `cbs/wijken-en-buurten-2025` dataset has `gemeenten`, `wijken` and `buurten`
layers. We take the boundary of Amsterdam from the `gemeenten` layer, then read
the `buurten` (neighborhoods) inside it.

```{r read}
gemeenten <- pdok_read("cbs/wijken-en-buurten-2025", "gemeenten")
amsterdam <- filter(gemeenten, gemeentenaam == "Amsterdam")

buurten <- pdok_read(
  "cbs/wijken-en-buurten-2025", "buurten",
  filter_by = amsterdam, predicate = "within"
)
nrow(buurten)
```

## Handle the "no data" code

CBS marks a figure that is unknown or suppressed (too few cases, or a
non-residential area) with a large negative sentinel such as `-99997`. We turn
those into `NA` so they do not distort the map.

```{r clean}
buurten <- buurten |>
  mutate(bevolkingsdichtheid_inwoners_per_km2 =
           if_else(bevolkingsdichtheid_inwoners_per_km2 < 0, NA_real_,
                   bevolkingsdichtheid_inwoners_per_km2))
summary(buurten$bevolkingsdichtheid_inwoners_per_km2)
```

The value is the number of inhabitants per square kilometer.

## Draw the choropleth

A quantile classification keeps each color class roughly equal in size, which
works well for a skewed variable like density. Neighborhoods without a value
(water, parks, industry) are drawn in gray.

```{r map-static}
tmap_mode("plot")

tm_basemap(pdok_basemap("grijs")) +
  tm_shape(buurten) +
  tm_polygons(
    fill = "bevolkingsdichtheid_inwoners_per_km2",
    fill.scale = tm_scale_intervals(n = 6, style = "quantile",
                                    values = "brewer.yl_or_rd"),
    fill.legend = tm_legend("Inhabitants per km2"),
    col = "white", lwd = 0.2, fill_alpha = 0.8
  ) +
  tm_title("Population density by neighborhood, Amsterdam (CBS 2025)") +
  tm_credits("Kaartgegevens © Kadaster")
```

The same map interactively — hover for the neighborhood name and click for the
value. A little transparency lets the basemap show through, so you keep your
bearings:

```{r map-interactive}
tmap_mode("view")

tm_basemap(pdok_basemap("grijs")) +
  tm_shape(buurten) +
  tm_polygons(
    fill = "bevolkingsdichtheid_inwoners_per_km2",
    fill.scale = tm_scale_intervals(n = 6, style = "quantile",
                                    values = "brewer.yl_or_rd"),
    fill.legend = tm_legend("Inhabitants per km2"),
    col = "white", lwd = 0.2, fill_alpha = 0.6,
    id = "buurtnaam",
    popup = tm_popup(vars = c("Inhabitants per km2" = "bevolkingsdichtheid_inwoners_per_km2"))
  ) +
  tm_credits("Kaartgegevens © Kadaster")
```

```{r reset, include = FALSE}
tmap_mode("plot")
```

To map a different figure, swap `bevolkingsdichtheid_inwoners_per_km2` for
another column — `aantal_inwoners`, `stedelijkheid_adressen_per_km2`, or
`gemiddelde_woningwaarde`, among hundreds of others (some figures appear a year
or two after the boundaries).

## Where to next

* [Filtering data by area](filtering-by-area.html) — the load-then-filter
  workflow.
* [Coordinate reference systems](coordinate-reference-systems.html) — working in
  RD New (EPSG:28992), the Dutch national grid.
* [PDOK basemaps](basemaps.html) — the gray background map used here, and the
  other styles.
