---
title: "Coordinate reference systems"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%"
)
```

Dutch geodata lives in a few different coordinate reference systems (CRS), and
mixing them up is a classic source of confusion. This article explains which
ones you will meet, how PDOK serves them, and how `pdokr` handles them.

```{r setup}
library(pdokr)
library(tmap)
```

## The CRSs you will meet

| EPSG | Name | Units | Where |
|------|------|-------|-------|
| 28992 | Amersfoort / RD New | meters | the Dutch national grid; most data is *stored* in this |
| 4326 | WGS 84 (CRS84, lon/lat) | degrees | the default for web/GeoJSON |
| 3857 | Web Mercator | meters | web map tiles |
| 4258 | ETRS89 (lon/lat) | degrees | the European reference frame |

## What PDOK offers per layer

`pdok_list_layers()` reports, for each layer, the CRSs it can deliver (`crs`) and
the CRS the data is stored in (`storage_crs`).

```{r layers}
layers <- pdok_list_layers("cbs/gebiedsindelingen")
gemeente <- layers[layers$layer == "gemeente_gegeneraliseerd", ]

gemeente$crs[[1]]
gemeente$storage_crs
```

So this layer is stored in RD New (28992) but can be served in several CRSs.

## What `pdok_read()` returns

Over the OGC API, PDOK serves GeoJSON, whose default is CRS84 (lon/lat). So by
default `pdok_read()` gives you lon/lat:

```{r default}
gemeenten <- pdok_read(
  "cbs/gebiedsindelingen", "gemeente_gegeneraliseerd",
  datetime = 2025
)
sf::st_crs(gemeenten)$epsg
```

Ask for another CRS with the `crs` argument. `pdokr` does this transformation
client-side with `sf::st_transform()`, so it works for any target CRS — even one
the service does not offer.

```{r rdnew}
gemeenten_rd <- pdok_read(
  "cbs/gebiedsindelingen", "gemeente_gegeneraliseerd",
  datetime = 2025, crs = 28992
)
sf::st_crs(gemeenten_rd)$epsg
```

The coordinates differ accordingly — degrees versus meters:

```{r bbox}
sf::st_bbox(gemeenten)
sf::st_bbox(gemeenten_rd)
```

## Why it matters for maps

The same data plotted in lon/lat and in RD New is *not* the same picture. Lon/lat
is unprojected, so at Dutch latitudes (~52° N) the country looks horizontally
stretched; RD New is an accurate planar projection.

The lon/lat graticule (the grid of meridians and parallels) makes the difference
obvious: it is a perfect rectangular grid in lon/lat, but the *same* lines run
slightly skewed in RD New, because that projection is gently rotated relative to
north.

```{r map-lonlat}
tmap_mode("plot")
tm_shape(gemeenten) +
  tm_polygons(fill = "grey95", col = "grey50", lwd = 0.3) +
  tm_graticules(col = "steelblue", lwd = 0.5) +
  tm_title("Lon/lat (EPSG:4326)")
```

```{r map-rdnew}
tm_shape(gemeenten_rd) +
  tm_polygons(fill = "grey95", col = "grey50", lwd = 0.3) +
  tm_graticules(col = "steelblue", lwd = 0.5) +
  tm_title("RD New (EPSG:28992)")
```

For analysis in meters (distances, areas, buffers) use RD New; for web maps,
lon/lat or Web Mercator.

## Two things `pdokr` handles for you

* **The `bbox` pre-filter is always CRS84.** Whatever CRS your `bbox` or
  `filter_by` geometry is in, `pdokr` converts its extent to lon/lat before
  sending it to the server, because that is what the OGC `bbox` parameter
  expects.
* **Axis order.** "CRS84" means longitude/latitude order, while EPSG:4326 is
  formally latitude/longitude. `sf` consistently stores coordinates as
  x = longitude, y = latitude, and `pdokr` follows that, so you do not have to
  think about axis swaps.

## Where to next

* [Getting started](getting-started.html) — the basics of reading and mapping a
  layer.
* [Working with PDOK by hand](pdok-by-hand.html) — what `pdokr` does under the
  hood, CRS handling included.
