---
title: "Combining with external data: school locations (DUO)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

PDOK is rarely the only source you need. A common pattern is to take a table
from another open API, turn it into spatial data, and combine it with
authoritative PDOK geometry. This article does that with
[DUO](https://duo.nl/)'s open data on schools — and shows that combining sources
often means a little data wrangling first.

```{r setup}
library(pdokr)
library(tmap)
library(sf)
library(jsonlite)
```

DUO publishes its data through a [CKAN open-data
API](https://onderwijsdata.duo.nl/). Every table is one resource; this small
helper fetches one as a data frame.

```{r helper}
read_duo <- function(resource_id) {
  url <- paste0(
    "https://onderwijsdata.duo.nl/api/3/action/datastore_search",
    "?resource_id=", resource_id, "&limit=50000"
  )
  fromJSON(url)$result$records
}
```

## 1. Fetch the education locations

The `onderwijslocaties` resource lists every location where education *may* take
place, each with coordinates and a `BAG_ID`.

```{r locations}
locations <- read_duo("a7e3f323-6e46-4dca-a834-369d9d520aa8")
nrow(locations)
```

There is a catch: these are *possible* education locations, not recognised
schools. A conference centre that hosts the odd exam, or a commercial training
firm, is in here too. Mapping them straight away would put dots where no school
exists.

## 2. Keep only the recognised schools

To find the schools recognised in law we combine two more DUO tables. The
`relaties...` resource links a location (`ONDERWIJSLOCATIECODE`) to a recognised
institution (`VESTIGINGSCODE`); the `vestigingserkenningen` resource describes
that institution — its name and the education law it falls under (`WET`). Both
tables are historical, so a row counts only when it has no `EINDDATUM`.

```{r join}
relations    <- read_duo("c18ec7dd-aa4e-4b51-997c-782955f1aa38")
institutions <- read_duo("01fd2a5f-40af-456f-864d-13265a51e5e2")

# keep only currently valid rows (no end date)
relations    <- relations[is.na(relations$EINDDATUM), ]
institutions <- institutions[is.na(institutions$EINDDATUM), ]
institutions <- institutions[!duplicated(institutions$VESTIGINGSCODE), ]

# translate the education-law code into a readable sector
sectors <- c(WPO = "Primary", WVO = "Secondary", WEC = "Special",
             WEB = "Vocational (MBO)", WHW = "Higher education")
institutions$sector <- sectors[institutions$WET]

# location -> recognised institution -> sector and name (one per location)
link <- merge(
  relations[, c("ONDERWIJSLOCATIECODE", "VESTIGINGSCODE")],
  institutions[, c("VESTIGINGSCODE", "sector", "VOLLEDIGE_NAAM")],
  by = "VESTIGINGSCODE"
)
link <- link[!duplicated(link$ONDERWIJSLOCATIECODE), ]
names(link)[names(link) == "VOLLEDIGE_NAAM"] <- "school"

# inner join drops every location without a recognised institution
schools <- merge(locations, link, by = "ONDERWIJSLOCATIECODE")
nrow(schools)
table(schools$sector)
```

Only the locations backed by a recognised institution remain — no training
venues, no exam halls — and each carries its school name and education sector.

One quirk worth knowing: a university or college is registered as a *single*
recognised institution, so only its official location appears here, not its
every building. The map below is therefore dominated by primary and secondary
schools, which register each site.

## 3. Make it spatial and combine with PDOK

The coordinates are WGS84 longitude/latitude, so we build an `sf` object and use
`pdok_filter_by()` with a municipal boundary to zoom in on Utrecht.

```{r spatial}
schools <- schools[!is.na(schools$GPS_LONGITUDE) & !is.na(schools$GPS_LATITUDE), ]
schools <- st_as_sf(
  schools, coords = c("GPS_LONGITUDE", "GPS_LATITUDE"), crs = 4326
)

gemeenten <- pdok_read(
  "cbs/gebiedsindelingen", "gemeente_gegeneraliseerd", datetime = 2025
)
utrecht <- gemeenten[gemeenten$statnaam == "Utrecht", ]

utrecht_schools <- pdok_filter_by(schools, utrecht, predicate = "within")
nrow(utrecht_schools)
```

```{r map-overview}
tmap_mode("plot")

tm_shape(utrecht) +
  tm_polygons(fill = "grey95", col = "grey60") +
  tm_shape(utrecht_schools) +
  tm_dots(
    fill = "sector", size = 0.5,
    fill.scale = tm_scale_categorical(values = "brewer.set2"),
    fill.legend = tm_legend("Sector")
  ) +
  tm_title("Recognised schools in Utrecht, by sector")
```

## 4. Map the school buildings

Each school sits in a building from the BAG. We read the `pand` (building) layer
for the historic centre, keep the footprints that contain a school, and carry
the school name and sector across with a spatial join.

```{r footprints}
wijken <- pdok_read(
  "cbs/gebiedsindelingen", "wijk_gegeneraliseerd", datetime = 2025,
  filter_by = utrecht, predicate = "within"
)
binnenstad <- wijken[grepl("Binnenstad", wijken$statnaam), ]
centre_schools <- pdok_filter_by(utrecht_schools, binnenstad, predicate = "within")

panden <- pdok_read("kadaster/bag", "pand", filter_by = binnenstad)
school_buildings <- st_filter(panden, centre_schools)
school_buildings <- st_join(
  school_buildings, centre_schools[, c("school", "sector", "STRAATNAAM")]
)
nrow(school_buildings)
```

The result is real school buildings, coloured by sector. Click one for the
school that uses it.

```{r map-interactive}
tmap_mode("view")

tm_basemap("CartoDB.Positron") +
  tm_shape(school_buildings) +
  tm_polygons(
    fill = "sector",
    fill.scale = tm_scale_categorical(values = "brewer.set2"),
    fill.legend = tm_legend("Sector"),
    col = "grey30", lwd = 0.5, id = "school",
    popup = tm_popup(vars = c("School" = "school", "Sector" = "sector",
                              "Street" = "STRAATNAAM"))
  )
```

```{r reset, include = FALSE}
tmap_mode("plot")
```

## Where to next

* [Filtering data by area](filtering-by-area.html) — more on `pdok_filter_by()`.
* [Mapping buildings by construction year](bag-buildings.html) — another BAG
  example, using the `pand` (building) layer.
