---
title: "Features and rasters"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

PDOK data comes in two fundamentally different shapes, and `pdokr` gives you
both. Most functions return **vector features** as `sf`; `pdok_ahn()` returns
**raster** data as a `terra` object. This article explains the difference, when
you reach for which, and how the two work together.

```{r setup}
library(pdokr)
library(sf)
library(terra)
library(tmap)
```

## Features: discrete objects

A **feature** is a *thing* with a geometry (point, line, or polygon) and
attributes. A building, a road, a municipality. `pdok_read()` returns features
as an `sf` object — one row per object. (We ask for RD New coordinates,
EPSG:28992, so they line up with the AHN raster later on.)

```{r features}
area <- st_as_sfc(st_bbox(
  c(xmin = 136300, ymin = 456000, xmax = 136600, ymax = 456300), crs = 28992
))
buildings <- pdok_read("kadaster/bag", "pand", filter_by = area, crs = 28992)

buildings[, "bouwjaar"]
```

Each building is a polygon you can select, filter, and query by attribute.

## Rasters: a grid of values

A **raster** is a *grid of cells*, each holding a value — a continuous surface
rather than a set of objects. AHN, the national height model, is a raster: every
0.5 × 0.5 m cell holds an elevation in metres (NAP).

We will look at a stretch of North Sea coastal dunes, where the relief is
dramatic. First, so you know *where* we are looking — the red square marks the
spot (zoom out to place it on the coast):

```{r locator}
dunes <- st_as_sfc(st_bbox(
  c(xmin = 99000, ymin = 487500, xmax = 100500, ymax = 489000), crs = 28992
))

tmap_mode("view")
tm_basemap(pdok_basemap("grijs")) +
  tm_shape(dunes, bbox = st_bbox(
    c(xmin = 88000, ymin = 478000, xmax = 112000, ymax = 500000), crs = 28992
  )) +
  tm_borders(col = "red", lwd = 2) +
  tm_credits("Kaartgegevens © Kadaster")
```

```{r reset, include = FALSE}
tmap_mode("plot")
```

`pdok_ahn()` returns the elevation there as a `terra` `SpatRaster`:

```{r raster}
elevation <- pdok_ahn(dunes, model = "dtm")
elevation
```

```{r raster-plot}
plot(elevation, main = "Terrain height (m NAP) — coastal dunes")
```

There are no "objects" here — just a height value everywhere.

## When do you use which?

| | Features (`sf`) | Raster (`terra`) |
|---|---|---|
| Data model | discrete objects + attributes | grid of cells with values |
| Good for | things with boundaries & identities | continuous phenomena |
| Examples | buildings, parcels, boundaries, roads | elevation, temperature, imagery, land cover |
| Query | "which buildings are …?" | "what is the height at …?" |
| pdokr | `pdok_read()`, `pdok_geocode()`, `pdok_filter_by()` | `pdok_ahn()` |

## Combining the two

The real power is combining them: `terra::extract()` reads raster values onto
features. Because both our buildings and the AHN raster are in RD New, we can
extract directly — for each building, the median surface height within its
footprint (its roof top above NAP):

```{r combine}
surface <- pdok_ahn(area, model = "dsm")

buildings$top_m <- terra::extract(surface, buildings, fun = median, na.rm = TRUE)$dsm

buildings[, "top_m"]
```

The same idea with the *surface* minus the *terrain* model gives building
*heights* — the basis for a 3D map (see below).

## Reading the landscape

Because AHN is so detailed, a little raster maths reveals the landscape. A
*hillshade* (a simulated sun angle over the terrain) turns the dune heights into
a relief image — impossible with features alone:

```{r hillshade}
slope  <- terrain(elevation, "slope", unit = "radians")
aspect <- terrain(elevation, "aspect", unit = "radians")
hillshade <- shade(slope, aspect)

plot(hillshade, col = grey(0:100 / 100), legend = FALSE,
     main = "Hillshade — the dunes revealed")
```

AHN this way exposes dunes, dikes, ancient terpen and old riverbeds across the
otherwise flat Netherlands.

## Where to next

* [3D buildings from BAG and AHN](3d-buildings.html) — combine footprints with
  heights into a 3D map.
* [Getting started](getting-started.html) — the vector basics.
