---
title: "Filtering data by area"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

A very common task is *"give me this layer, but only within that area"* — all
national parks in a province, all wells in a water-authority district, every
building in a municipality. PDOK cannot do this directly: its services can
pre-filter by a rectangle (a bounding box), but not by an arbitrary polygon.
`pdokr` bridges that gap, and `pdok_read()` makes it a one-liner.

```{r setup}
library(pdokr)
library(tmap)
library(dplyr)
```

## Get a filter geometry

The area you filter by is just an `sf` polygon. The most common source is the
CBS administrative boundaries (`"cbs/gebiedsindelingen"`). Here we take the
province of Fryslân.

```{r geometry}
provinces <- pdok_read(
  "cbs/gebiedsindelingen", "provincie_gegeneraliseerd",
  datetime = 2025
)
fryslan <- filter(provinces, statnaam == "Fryslân")
```

Any polygon works, though: a nature reserve, a water-authority area, a polygon
you drew yourself, or a result from `pdok_geocode()` (see the last section).

## Filter another layer to that area

Pass the filter geometry to `pdok_read()` as `filter_by`. It uses the polygon's
bounding box as a cheap server-side pre-filter, then clips the result to the
exact shape. Here we keep only the national parks that lie within Fryslân.

```{r filter}
parks <- pdok_read(
  "rvo/nationale-parken-geharmoniseerd", "protectedsite",
  filter_by = fryslan
)
parks$text
```

A static map of the province with its national parks:

```{r map-static}
tmap_mode("plot")

tm_shape(fryslan) +
  tm_borders(col = "grey40") +
  tm_shape(parks) +
  tm_polygons(fill = "text", fill.legend = tm_legend("National park")) +
  tm_title("National parks in Fryslân")
```

The same in interactive mode — zoom in and click a park:

```{r map-interactive}
tmap_mode("view")

tm_basemap(pdok_basemap("grijs")) +
  tm_shape(parks) +
  tm_polygons(fill = "text", fill_alpha = 0.6, id = "text",
              fill.legend = tm_legend("National park")) +
  tm_credits("Kaartgegevens © Kadaster")
```

```{r reset, include = FALSE}
tmap_mode("plot")
```

## What happens under the hood

`filter_by` folds a two-stage workflow into one call. You can also run the two
stages yourself, which is useful when you want to reuse the pre-filtered data or
apply a different spatial predicate.

```{r two-stage}
# Stage 1: cheap server-side pre-filter by the bounding box of Fryslân
parks_in_bbox <- pdok_read(
  "rvo/nationale-parken-geharmoniseerd", "protectedsite",
  bbox = fryslan
)

# Stage 2: exact client-side clip to the polygon
pdok_filter_by(parks_in_bbox, fryslan)$text
```

`pdok_filter_by()` reconciles coordinate reference systems for you and accepts a
`predicate` that sets *how* a feature must relate to the area to be kept:
`"intersects"` (the default — touching the area in any way), `"within"` (lying
entirely inside it), `"contains"`, `"disjoint"` (everything outside), and the
more specialized `"overlaps"`, `"touches"`, `"crosses"`, `"covers"` and
`"covered_by"`. Each maps to the matching `sf` function `sf::st_<predicate>()`;
see [`?sf::geos_binary_pred`](https://r-spatial.github.io/sf/reference/geos_binary_pred.html)
for the precise definition of each. The plain-`sf` equivalent, once both objects
share a CRS, is:

```{r plain-sf}
fryslan_ll <- sf::st_transform(fryslan, sf::st_crs(parks_in_bbox))
parks_in_bbox[fryslan_ll, , op = sf::st_intersects]$text
```

## From an address to its area

Because `filter_by` also accepts a point, you can answer *"which municipality is
this address in?"* Geocode the address with `pdok_geocode()`, then filter the
municipalities by that point.

```{r address}
address <- pdok_geocode("Tweebaksmarkt 52, Leeuwarden")

pdok_read(
  "cbs/gebiedsindelingen", "gemeente_gegeneraliseerd",
  datetime = 2025, filter_by = address
)$statnaam
```

The same idea scales up: geocode a place as a polygon (for example
`pdok_geocode("Fryslân", type = "provincie")`) and use it directly as
`filter_by`.

## Where to next

* [Geocoding](geocoding.html) — turn a place name into the boundary you filter
  by.
* [Mapping buildings by construction year](bag-buildings.html) — read a BAG
  layer inside an area.
* [Thematic maps from CBS statistics](cbs-statistics.html) — a choropleth for
  one municipality's neighborhoods.
* [PDOK basemaps](basemaps.html) — the gray background map used here, and the
  other styles.
