---
title: "Geocoding: from names to locations"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

`pdok_geocode()` turns a name — an address, a postcode, a street, a town, a
municipality, a province — into spatial data, using the PDOK
[Locatieserver](https://www.pdok.nl/). The result is an `sf` object, so it plugs
straight into the rest of `pdokr`: most usefully, a geocoded boundary can be
handed to the `filter_by` argument of `pdok_read()`.

```{r setup}
library(pdokr)
library(tmap)
```

## 1. A name to a point

In its simplest form you pass a query and get back the single best match.

```{r address}
home <- pdok_geocode("Domplein 1, Utrecht")
home[, c("weergavenaam", "type", "score")]
```

It is an ordinary `sf` object with a point geometry, ready to map.

```{r address-map}
tmap_mode("view")

tm_basemap("CartoDB.Positron") +
  tm_shape(home) +
  tm_dots(fill = "#E8631C", size = 1)
```

```{r reset1, include = FALSE}
tmap_mode("plot")
```

## 2. Search levels: the `type` argument

The Locatieserver does not only know addresses. Every result has a `type`, and
the geometry you get back depends on it — a point for pinpoint locations, a line
for a road, a polygon for an area.

| `type` | Geometry | What it is |
|--------|----------|------------|
| `adres` | point | a specific address |
| `postcode` | point | the centre of a postcode |
| `hectometerpaal` | point | a motorway distance marker |
| `appartementsrecht` | point | an apartment right |
| `weg` | line | a (named) road |
| `buurt`, `wijk` | polygon | a neighbourhood or district |
| `woonplaats` | polygon | a town or city |
| `gemeente` | polygon | a municipality |
| `provincie` | polygon | a province |
| `perceel` | polygon | a cadastral parcel |

By default `pdok_geocode()` returns the best match of *any* type, ranked by the
service's relevance `score`. That is convenient, but a single name often exists
at several levels — "Utrecht" is a municipality, a province, *and* a town:

```{r ambiguous}
pdok_geocode("Utrecht", limit = 5)[, c("weergavenaam", "type", "score")]
```

Pass `type` to pin down exactly which level you mean. The same query then
returns very different geometry:

```{r levels}
point <- pdok_geocode("Domplein 1, Utrecht")          # adres  -> point
road  <- pdok_geocode("Oudegracht, Utrecht", type = "weg")   # weg    -> line
area  <- pdok_geocode("Binnenstad, Utrecht", type = "wijk")  # wijk   -> polygon

sf::st_geometry_type(point)
sf::st_geometry_type(road)
sf::st_geometry_type(area)
```

Seen together, the three levels nest neatly — the address sits on the road,
which lies inside the district.

```{r levels-map}
tm_shape(area) +
  tm_polygons(fill = "#F7F3EC", col = "grey50") +
  tm_shape(road) +
  tm_lines(col = "#1f78b4", lwd = 2) +
  tm_shape(point) +
  tm_dots(fill = "#E8631C", size = 0.6) +
  tm_title("Three search levels: district, road, address")
```

## 3. Combine with `pdok_filter_by()`

This is where geocoding earns its place in a workflow. Many tasks start with
"the data inside *this* area" — and `pdok_geocode()` gives you that area as a
boundary without having to find and load an administrative layer first. A
geocoded `gemeente`, `wijk` or `provincie` polygon drops straight into
`filter_by`.

```{r combine}
boundary <- pdok_geocode("De Bilt", type = "gemeente")

# read the neighbourhoods that meet the boundary, then keep those whose centre
# lies inside it. A geocoded boundary is generalised, so it clips edge
# neighbourhoods unevenly; `predicate = "within"` alone would drop some, while
# `"intersects"` pulls in slivers of neighbouring municipalities.
buurten <- pdok_read(
  "cbs/gebiedsindelingen", "buurt_gegeneraliseerd", datetime = 2025,
  filter_by = boundary, predicate = "intersects"
)
centres <- suppressWarnings(sf::st_centroid(buurten))
buurten <- buurten[lengths(sf::st_within(centres, boundary)) > 0, ]
nrow(buurten)
```

The geocoded boundary did the filtering; the result is every neighbourhood
inside the municipality of De Bilt. On an interactive map you can see where that
is and hover for the neighbourhood names:

```{r combine-map}
tmap_mode("view")

tm_basemap("CartoDB.Positron") +
  tm_shape(buurten) +
  tm_polygons(
    fill = "statnaam",
    fill.scale = tm_scale_categorical(values = "brewer.set3"),
    fill.legend = tm_legend(show = FALSE),
    col = "white", fill_alpha = 0.6, id = "statnaam"
  )
```

```{r reset3, include = FALSE}
tmap_mode("plot")
```

The same pattern works for any layer: geocode the area you care about, then read
the buildings, parcels or statistics within it.

## Where to next

* [Filtering data by area](filtering-by-area.html) — more on `pdok_filter_by()`.
* [Mapping buildings by construction year](bag-buildings.html) — read a BAG
  layer inside an area.
