---
title: "Geocoding: from names to locations"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

`pdok_geocode()` turns a name — an address, a postcode, a street, a town, a
municipality, a province — into spatial data, using the PDOK
[Locatieserver](https://www.pdok.nl/). The result is an `sf` object, so it plugs
straight into the rest of `pdokr`: most usefully, a geocoded boundary can be
handed to the `filter_by` argument of `pdok_read()`.

```{r setup}
library(pdokr)
library(tmap)
library(dplyr)
```

## A name to a point

In its simplest form you pass a query and get back the single best match.

```{r address}
home <- pdok_geocode("Domplein 1, Utrecht")
home |> select(weergavenaam, type, score)
```

It is an ordinary `sf` object with a point geometry, ready to map.

```{r address-map}
tmap_mode("view")

tm_basemap(pdok_basemap("grijs")) +
  tm_shape(home) +
  tm_dots(fill = "#E8631C", size = 1) +
  tm_credits("Kaartgegevens © Kadaster")
```

```{r reset1, include = FALSE}
tmap_mode("plot")
```

## Search levels: the `type` argument

The Locatieserver does not only know addresses. Every result has a `type`, and
the geometry you get back depends on it — a point for pinpoint locations, a line
for a road, a polygon for an area.

| `type` | Geometry | What it is |
|--------|----------|------------|
| `adres` | point | a specific address |
| `postcode` | point | the center of a postcode |
| `hectometerpaal` | point | a motorway distance marker |
| `appartementsrecht` | point | an apartment right |
| `weg` | line | a (named) road |
| `buurt`, `wijk` | polygon | a neighborhood or district |
| `woonplaats` | polygon | a town or city |
| `gemeente` | polygon | a municipality |
| `provincie` | polygon | a province |
| `perceel` | polygon | a cadastral parcel |

By default `pdok_geocode()` returns the best match of *any* type, ranked by the
service's relevance `score`. That is convenient, but a single name often exists
at several levels — "Utrecht" is a municipality, a province, *and* a town:

```{r ambiguous}
pdok_geocode("Utrecht", limit = 5) |>
  select(weergavenaam, type, score)
```

Pass `type` to pin down exactly which level you mean. The same query then
returns very different geometry:

```{r levels}
point <- pdok_geocode("Domplein 1, Utrecht")          # adres  -> point
road  <- pdok_geocode("Oudegracht, Utrecht", type = "weg")   # weg    -> line
area  <- pdok_geocode("Binnenstad, Utrecht", type = "wijk")  # wijk   -> polygon

sf::st_geometry_type(point)
sf::st_geometry_type(road)
sf::st_geometry_type(area)
```

Seen together, the three levels nest neatly — the address sits on the road,
which lies inside the district.

```{r levels-map}
tm_shape(area) +
  tm_polygons(fill = "#F7F3EC", col = "grey50") +
  tm_shape(road) +
  tm_lines(col = "#1f78b4", lwd = 2) +
  tm_shape(point) +
  tm_dots(fill = "#E8631C", size = 0.6) +
  tm_title("Three search levels: district, road, address")
```

## Combine with `pdok_filter_by()`

This is where geocoding earns its place in a workflow. Many tasks start with
"the data inside *this* area" — and `pdok_geocode()` gives you that area as a
boundary without having to find and load an administrative layer first. A
geocoded `gemeente`, `wijk` or `provincie` polygon drops straight into
`filter_by`.

```{r combine}
boundary <- pdok_geocode("De Bilt", type = "gemeente")

# read the neighborhoods that meet the boundary, then keep those whose center
# lies inside it. A geocoded boundary is generalized, so it clips edge
# neighborhoods unevenly; `predicate = "within"` alone would drop some, while
# `"intersects"` pulls in slivers of neighboring municipalities.
buurten <- pdok_read(
  "cbs/gebiedsindelingen", "buurt_gegeneraliseerd", datetime = 2025,
  filter_by = boundary, predicate = "intersects"
)
centers <- suppressWarnings(sf::st_centroid(buurten))
buurten <- filter(buurten, lengths(sf::st_within(centers, boundary)) > 0)
nrow(buurten)
```

The geocoded boundary did the filtering; the result is every neighborhood
inside the municipality of De Bilt. On an interactive map you can see where that
is and hover for the neighborhood names:

```{r combine-map}
tmap_mode("view")

tm_basemap(pdok_basemap("grijs")) +
  tm_shape(buurten) +
  tm_polygons(
    fill = "statnaam",
    fill.scale = tm_scale_categorical(values = "brewer.set3"),
    fill.legend = tm_legend(show = FALSE),
    col = "white", fill_alpha = 0.6, id = "statnaam"
  ) +
  tm_credits("Kaartgegevens © Kadaster")
```

```{r reset3, include = FALSE}
tmap_mode("plot")
```

The same pattern works for any layer: geocode the area you care about, then read
the buildings, parcels or statistics within it.

## Geocoding a whole column

`pdok_geocode()` takes a *vector* of queries, so you can geocode a column of
addresses in one call. Every result row carries a `query` column identifying the
input it came from, which keeps the output aligned with the input — even when a
query returns several candidates, or none.

Free search ranks matches across *all* levels — the exact address, but also the
street or place it sits on — and returns the best-ranked one per query. So keep
an eye on the `type` column: below, two queries resolve to the street (`weg`)
rather than the house number, which is why their geometry is a line, not a point.

```{r batch}
addresses <- c("Domplein 1, Utrecht", "Coolsingel 40, Rotterdam",
               "Vrijthof 1, Maastricht")

pdok_geocode(addresses)[, c("query", "type", "weergavenaam")]
```

Pin the level with `type` when you know what you want. `type = "adres"` returns
the exact addresses, all as points:

```{r batch-adres}
pdok_geocode(addresses, type = "adres")[, c("query", "type", "weergavenaam")]
```

## The other direction: reverse geocoding

Sometimes you have coordinates — GPS points, sensor locations, a spot on a map —
and want to know *what is there*. `pdok_reverse_geocode()` takes `sf` points and
returns the nearest address, road or place, with an `afstand` column giving the
distance in meters. A `point_id` column maps each match back to its input point.

```{r reverse}
points <- sf::st_sfc(
  sf::st_point(c(5.121, 52.090)),   # central Utrecht
  sf::st_point(c(4.479, 51.923)),   # central Rotterdam
  crs = 4326
)

pdok_reverse_geocode(points)[, c("point_id", "type", "weergavenaam", "afstand")]
```

By default it returns the nearest match of any type — usually an address. Use
`type` to choose the level: `type = "weg"` for the nearest road, or
`type = "gemeente"` for the municipality the point falls in (distance 0, since
the point is inside it).

The input can be in any CRS — RD New points work just as well; they are
converted to lon/lat for the query automatically.

## Where to next

* [Filtering data by area](filtering-by-area.html) — more on `pdok_filter_by()`.
* [Mapping buildings by construction year](bag-buildings.html) — read a BAG
  layer inside an area.
* [PDOK basemaps](basemaps.html) — the gray background map used here, and the
  other styles.
