---
title: "Getting started"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

`pdokr` helps you work with open geographic data from
[PDOK](https://www.pdok.nl/), the national geodata platform of the Netherlands.
A typical session has three steps: find a dataset and its layers, load a layer
as an [`sf`](https://r-spatial.github.io/sf/) object, and take it from there.

```{r setup}
library(pdokr)
```

## Find a dataset

PDOK serves many datasets. Search them by a partial, case-insensitive term with
`pdok_search_datasets()`, or list them all with `pdok_list_datasets()`.

```{r search}
pdok_search_datasets("gebiedsindelingen")
```

The `id` column is what the rest of the package uses. Here we will use the CBS
administrative boundaries, `"cbs/gebiedsindelingen"`.

## List the layers in a dataset

A dataset contains one or more layers. List them with `pdok_list_layers()` (or
filter them with `pdok_search_layers()`).

```{r layers}
pdok_list_layers("cbs/gebiedsindelingen")
```

Each row carries its `dataset`, the `layer` id you pass to `pdok_read()`, the
period the layer covers (`start_date` / `end_date`), and the coordinate
reference systems it offers.

## Load a layer

`pdok_read()` loads a layer as an `sf` object, handling pagination for you. The
CBS boundaries hold every year in one layer, so we select 2025 with `datetime`.

```{r read}
gemeenten <- pdok_read(
  "cbs/gebiedsindelingen", "gemeente_gegeneraliseerd",
  datetime = 2025
)
gemeenten
```

## Make a map

The result is a plain `sf` object, so any mapping package works. We use
[`tmap`](https://r-tmap.github.io/tmap/). A static overview:

```{r map-static}
library(tmap)
tmap_mode("plot")

tm_shape(gemeenten) +
  tm_polygons(fill = "grey95", col = "grey40", lwd = 0.4) +
  tm_title("Dutch municipalities, 2024")
```

The same code in `tmap`'s interactive mode gives a zoomable, pannable map you
can hover over — try it:

```{r map-interactive}
tmap_mode("view")

tm_basemap(pdok_basemap("grijs")) +
  tm_shape(gemeenten) +
  tm_polygons(fill = "red", fill_alpha = 0, col = "red", lwd = 0.8,
              id = "statnaam") +
  tm_credits("Kaartgegevens © Kadaster")
```

```{r reset, include = FALSE}
tmap_mode("plot")
```

## Where to next

* [Filtering data by area](filtering-by-area.html) — filter a layer to a
  municipality, province, or any polygon, including the address-to-area workflow
  with `pdok_geocode()`.
* [Coordinate reference systems](coordinate-reference-systems.html) — how
  `pdokr` handles RD New and lon/lat.
* [Working with PDOK by hand](pdok-by-hand.html) — the raw OGC API and WFS
  requests the package automates.
* [PDOK basemaps](basemaps.html) — the grey background map used here, and the
  other styles.
