---
title: "Working with PDOK by hand"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

`pdokr` is a thin convenience layer over PDOK's open web services. This article
shows how to talk to those services directly with
[`httr2`](https://httr2.r-lib.org/) and [`sf`](https://r-spatial.github.io/sf/),
so you understand what the package automates and can drop down to raw requests
when you need to.

```{r setup}
library(httr2)
library(sf)
```

## The dataset index

Every OGC API dataset is listed at `https://api.pdok.nl/index.json`.

```{r index}
index <- request("https://api.pdok.nl/index.json") |>
  req_perform() |>
  resp_body_json()

length(index$apis)
index$apis[[1]]$links[[1]]$href
```

`pdok_list_datasets()` fetches and tidies exactly this.

## The layers in a dataset

Each dataset exposes its layers at `{base}/collections?f=json`.

```{r collections}
collections <- request("https://api.pdok.nl/cbs/gebiedsindelingen/ogc/v1/collections") |>
  req_url_query(f = "json") |>
  req_perform() |>
  resp_body_json()

vapply(collections$collections, function(x) x$id, character(1))[1:6]
```

That is what `pdok_list_layers()` parses.

## Reading features

Features come from `{base}/collections/{layer}/items` as GeoJSON. Add query
parameters for the format, a page size, and optional filters. Here we ask for the
2024 provinces.

```{r items}
url <- paste0(
  "https://api.pdok.nl/cbs/gebiedsindelingen/ogc/v1/",
  "collections/provincie_gegeneraliseerd/items"
)

resp <- request(url) |>
  req_url_query(
    f = "json",
    limit = 12,
    datetime = "2024-07-01T00:00:00Z"
  ) |>
  req_perform()

provinces <- read_sf(resp_body_string(resp))
provinces[, "statnaam"]
```

The CRS of the returned coordinates is announced in a response header (CRS84,
i.e. lon/lat, by default):

```{r crs}
resp_header(resp, "Content-Crs")
```

A bounding box pre-filter is just another query parameter,
`bbox = "xmin,ymin,xmax,ymax"` in CRS84.

## Pagination

The API does not return all features at once. Each response carries a `next`
link when more pages are available; you follow it until it disappears. (PDOK uses
cursor-based pagination, so you cannot jump to an offset.)

```{r pagination}
body <- resp_body_json(resp)
next_link <- Filter(function(l) l$rel == "next", body$links)
length(next_link) # 0 here: all 12 provinces fit on one page
```

`pdok_read()` runs this loop for you and binds the pages into one `sf` object.

## WFS, the older service

PDOK also offers an older Web Feature Service (WFS) for some datasets. `pdokr`
is an OGC API Features client and does not read WFS — almost all vector data
(including BAG) is now available over the OGC API. On the rare occasion you need
a WFS-only layer, GDAL can read it directly, optionally with a server-side
spatial filter:

```{r wfs, eval = FALSE}
# Read a WFS layer by hand (the layer's CRS is RD New / EPSG:28992):
buildings <- read_sf(
  "WFS:https://service.pdok.nl/some/dataset/wfs/v1_0",
  layer = "namespace:layer",
  wkt_filter = sf::st_as_text(area_in_rd_new)
)
```

## Putting it together

What we read by hand is an ordinary `sf` object, ready to map:

```{r map}
library(tmap)
tmap_mode("plot")

tm_shape(provinces) +
  tm_polygons(fill = "grey95", col = "grey40") +
  tm_title("Provinces, fetched by hand")
```

Everything above is what `pdokr` wraps: discovery, pagination, CRS handling,
and the GeoJSON-to-`sf` conversion, behind a handful of `pdok_*()` functions.
